SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2018-07-24 scipy-1.0/paper.tex
ENH: first draft of the `interpolate` subsection
Start rough draft of sparse matrices tech improvements section.

\author[1]{Pauli Virtanen}
\author[2,*]{Ralf Gommers}
\author[3,4]{Tyler Reddy}
\author[5]{Anne Archibald}
\author[6]{Andrew Nelson}
\author[7]{Charles Harris}
\author[8]{CJ Carey}
\author[9]{Denis Laxalde}
\author[10]{Eric Larson}
\author[11]{Eric Moore}
\author[12]{Eric Quintero}
\author[13]{Evgeni Burovski}
\author[14]{Jaime Fernández del Río}
\author[15]{Josef Perktold}
\author[16]{Josh Wilson}
\author[17]{Matthew Brett}
\author[18]{Nikolay Mayorov}
\author[19]{Warren Weckesser}
\author[20]{Matt Haberland}
\author[21]{Scott Sievert}
\author[22]{Yu Feng}
\author[23]{Antonio Horta Ribeiro}

\affil[1]{Affiliation, department, city, postcode, country}
\affil[2]{Affiliation, department, city, postcode, country}
\affil[2]{Affiliation, department, city, postcode, country}
\affil[3]{UC Berkeley Institute for Data Science,
          Berkeley, CA, 94720, USA}
\affil[4]{Los Alamos National Laboratory,
	  Theoretical Division 6,
          Los Alamos, NM, 87545, USA}
\affil[5]{Affiliation, department, city, postcode, country}
\affil[6]{Affiliation, department, city, postcode, country}
\affil[7]{Affiliation, department, city, postcode, country}
\affil[8]{Affiliation, department, city, postcode, country}
\affil[9]{Affiliation, department, city, postcode, country}
\affil[10]{Affiliation, department, city, postcode, country}
\affil[11]{Affiliation, department, city, postcode, country}
\affil[12]{Affiliation, department, city, postcode, country}
\affil[13]{Affiliation, department, city, postcode, country}
\affil[14]{Affiliation, department, city, postcode, country}
\affil[15]{Affiliation, department, city, postcode, country}
\affil[16]{Affiliation, department, city, postcode, country}
\affil[17]{Affiliation, department, city, postcode, country}
\affil[18]{Affiliation, department, city, postcode, country}
\affil[19]{Affiliation, department, city, postcode, country}
\affil[20]{Affiliation, department, city, postcode, country}
\affil[21]{Affiliation, department, city, postcode, country}
\affil[22]{Affiliation, department, city, postcode, country}
\affil[23]{Affiliation, department, city, postcode, country}

@@

\textbf{Project goals and scope}

@@

\subsection*{API and ABI evolution}
In general, we encourage changes that improve clarity in the API of the library,
but strongly discourage breaking backwards compatibility, given our position near
the base of the scientific Python computing stack.

@@

\subsection*{Data structures}
\textbf{cKDTree}
\input{ckdtree}
\textbf{Sparse matrices}

@@

@@

\subsection*{Statistical distributions}
Version 1.0 of SciPy saw the addition of the energy distance algorithm in
\texttt{scipy.stats.energy\_distance}. Energy distance is used in a varity of applications ranging
from genetic analysis to multivariate independence. We also added the
first Wasserstein distance algorithm to the library in
\texttt{scipy.stats.wasserstein\_distance}. The Wasserstein distance is often
used in color digital image analysis.

We now also provide the cumulative distribution function of the multivariate
normal distribution, \\\texttt{scipy.stats.multivariate\_normal}, via the newly
associated \texttt{logcdf} and \texttt{cdf} methods.

Other recent additions to \texttt{scipy.stats} include the Argus distribution
in \texttt{scipy.stats.argus} (used in particle physics), a class for
generating a distribution from a histogram in
\texttt{scipy.stats.rv\_histogram}, an implementation of the multinomial
distribution in \texttt{scipy.stats.multinomial} (used for example in natural
language processing), a weighted Kendall's tau statistic algorithm in
\texttt{scipy.stats.weightedtau}, interquantile range handling in
\texttt{scipy.stats.iqr}, the Burr type XII distribution is available in
\texttt{scipy.stats.burr12}, the skew normal distribution has been implemented 
as \texttt{scipy.stats.skewnorm}, a trepzoidal continuous random variable
has been added in \texttt{scipy.stats.trapz}, and the matrix normal distribution
is now available in \texttt{scipy.stats.matrix\_normal}.

From a design standpoint, many functions in \texttt{scipy.stats} now support
a \texttt{nan\_policy} keyword to allow user-level control over the algorithmic
decision to propagate, omit, or raise an error in the case that \texttt{NaN}
values are encountered.

@@

While \texttt{scipy.sparse} is mature, there are many data
formats and the community continues to move forward here so we will aim
to take advantage of their improvements\cite{Sparse_Arrays} 
while preserving backwards compatibility to the extent that it is possible.
\texttt{sparse.linalg} may benefit from wrappers for PROPACK 
for faster sparse SVD computation.

@@

\textbf{Maintainers and contributors}

@@

% need some better usage stats sources?
% SciPy stuff from njs blog: https://docs.google.com/spreadsheets/d/1lOLvSF0up4eZyv2ugZi-TM_GIs3gPCkGNIDKXS1Y3w4/edit#gid=406031940
From the user side, there have been at least 15 million downloads
of the SciPy manylinux wheels between the start of 2016 and May of 2018
(njs blog). The recommended source for citing SciPy is simply a reference
to the website / library itself\cite{SciPylib}, and has been cited $>3000$
times. Some of the most prominent citing articles include the IPython
publication\cite{PER-GRA:2007}, a heavily-cited Nature Biotechnology
article\cite{pmid17287757}, a dark
matter annihilitation study in the field of
cosmology\cite{1475-7516-2012-08-007}, and many of the
other major downstream software libraries in the Python scientific
computing stack.

\textbf{Future development}.
\textit{This section should include key issues: sparse arrays, ndimage pixel vs point, splines, fftpack vs. np.fft and linalg vs. np.linalg, under-maintained submodules.}

>>>>

\author[1]{Pauli Virtanen}
\author[2,*]{Ralf Gommers}
\author[3,4]{Tyler Reddy}
\author[5]{Anne Archibald}
\author[6]{Andrew Nelson}
\author[7]{Charles Harris}
\author[8]{CJ Carey}
\author[9]{Denis Laxalde}
\author[10]{Eric Larson}
\author[11]{Eric Moore}
\author[12]{Eric Quintero}
\author[13]{Evgeni Burovski}
\author[14]{Jaime Fernández del Río}
\author[15]{Josef Perktold}
\author[16]{Josh Wilson}
\author[17]{Matthew Brett}
\author[18]{Nikolay Mayorov}
\author[19]{Warren Weckesser}
\author[20]{Matt Haberland}
\author[21]{Scott Sievert}
\author[22]{Yu Feng}
\author[23]{Antonio Horta Ribeiro}
\author[24]{Ian Henriksen}

\affil[1]{Affiliation, department, city, postcode, country}
\affil[2]{Affiliation, department, city, postcode, country}
\affil[2]{Affiliation, department, city, postcode, country}
\affil[3]{UC Berkeley Institute for Data Science,
          Berkeley, CA, 94720, USA}
\affil[4]{Los Alamos National Laboratory,
	  Theoretical Division 6,
          Los Alamos, NM, 87545, USA}
\affil[5]{Affiliation, department, city, postcode, country}
\affil[6]{Affiliation, department, city, postcode, country}
\affil[7]{Affiliation, department, city, postcode, country}
\affil[8]{Affiliation, department, city, postcode, country}
\affil[9]{Affiliation, department, city, postcode, country}
\affil[10]{Affiliation, department, city, postcode, country}
\affil[11]{Affiliation, department, city, postcode, country}
\affil[12]{Affiliation, department, city, postcode, country}
\affil[13]{Affiliation, department, city, postcode, country}
\affil[14]{Affiliation, department, city, postcode, country}
\affil[15]{Affiliation, department, city, postcode, country}
\affil[16]{Affiliation, department, city, postcode, country}
\affil[17]{Affiliation, department, city, postcode, country}
\affil[18]{Affiliation, department, city, postcode, country}
\affil[19]{Affiliation, department, city, postcode, country}
\affil[20]{Affiliation, department, city, postcode, country}
\affil[21]{Affiliation, department, city, postcode, country}
\affil[22]{Affiliation, department, city, postcode, country}
\affil[23]{Affiliation, department, city, postcode, country}
\affil[24]{University of Texas at Austin,
           Instite for Computational Engineering and Sciences,
	   Austin, TX, 78712, USA}

@@

\textbf{Project scope}
SciPy aims to provide fundamental algorithms for scientific computing. The
breadth of its scope was derived from the Guide to Available Mathematical
Software classification system (GAMS\cite{boisvert1991guide}). In areas
that move relatively slowly, e.g. linear algebra, SciPy aims to provide
complete coverage. In other areas it aims to provide fundamental building
blocks while interacting well with other packages specialized in that area.
For example, SciPy provides what one expect to find in a
statistics textbook (probability distributions, hypothesis tests, frequency
statistics, correlation functions, and more), while Statsmodels provides
more advanced statistical estimators and inference methods, scikit-learn
covers machine learning, PyMC3, emcee and PyStan cover Bayesian statistics
and probabilistic modelling.

We use the following criteria to determine whether or not to include new
functionality in SciPy:
\begin{itemize}
    \item The algorithm is of relevance to multiple fields of science.
    \item The algorithm is demonstrably important.  E.g., it is classic
    enough to be included in textbooks, or it is based on a peer-reviewed article
    which has a significant number of citations.
\end{itemize}

In terms of software systems and architecture, SciPy's scope matches NumPy's:
algorithms for in-memory computing on single machines, with support for a wide
range of data types and process architectures. Distributed computing and support
for graphics processing units (GPUs) is explicitly out of scope.

@@

\subsection*{API and ABI evolution}
The application programming interface (API) for SciPy consists of
approximately 1500 functions and classes.
Our policy for evolving the API over time is that new functionality can be added,
while removing or changing existing functionality can only be done if the benefits
of that exceeds the (often significant) costs to users, \textit{and} only
after giving clear deprecation warnings to those users for at least one year.
In general, we encourage changes that improve clarity in the API of the
library but strongly discourage breaking backwards compatibility\footnote{
    Backwards compatibility means that new versions of SciPy will keep user
    code written for an older version working.}
given our position near the base of the scientific Python computing stack.

In addition to the Python API, SciPy has C and Cython interfaces in a number
of submodules. Therefore, we have to also consider the application binary
interface (ABI). This ABI has been stable for a long time, and we aim to
evolve it only in a backwards compatible way.

@@

\subsection*{Data structures}
\textbf{cKDTree}
\input{ckdtree}
\textbf{Sparse matrices}
Over the last three years our sparse matrix handling internals have
been rewritten and performance has been improved.

@@

When writing algorithms or wrappers in Cython\cite{behnel2011cython}, it is often convenient to call directly into the libraries that SciPy wraps rather than routing the call back through the Python interpreter.
SciPy currently includes Cython wrappers\cite{blas-lapack-wrappers-scipy-2015} for many BLAS and LAPACK\cite{LAPACK} routines and for many of the special functions provided in the \texttt{scipy.{\allowbreak}special} submodule.
These wrappers are available in the modules \texttt{scipy.{\allowbreak}linalg.{\allowbreak}cython\_blas}, \texttt{scipy.{\allowbreak}linalg.{\allowbreak}cython\_lapack}, and \texttt{scipy.{\allowbreak}special.{\allowbreak}cython\_special} respectively.
The low-level interfaces for Cython can also be used by third party libraries to get direct access to the functions in these libraries while avoiding the overhead of Python function calls{\textemdash}this can give performance gains of one or two orders of magnitude for many use cases.
Cython's cimport machinery allows Cython based modules to share C-level objects with other Cython modules without requiring that the third-party extensions link against any additional libraries.
Function pointers to wrappers for library routines can also be shared in this way, so that Cython modules using the interface provided by SciPy no longer have to worry about finding and linking against the original library.
This lets other extensions avoid the complexity of finding and using the correct libraries on the various possible Fortran compiler ABI and OS combinations.
The low-level Cython wrappers provide an OS and Fortran compiler independent interface for the libraries they wrap.

To help with both correctness and ease of maintenance, most of these low-level Cython wrappers are generated automatically.
The wrappers for BLAS and LAPACK are primarily generated using type information that is parsed from the BLAS and LAPACK source files using F2PY\cite{peterson2009f2py}, though a small number of routines use hand-written type signatures instead.
The input and output types of each routine are saved in a data file that is read at build time and used to generate the corresponding Cython wrapper files.
The wrappers in \texttt{scipy.{\allowbreak}special.{\allowbreak}cython\_special} are also generated from a data file containing type information for the wrapped routines.

Since SciPy can be built with LAPACK 3.4.0 or later, Cython wrappers are only provided for the routines that maintain a consistent interface across all supported LAPACK versions.
In other words, wrappers are provided for routines that are included in the earliest supported LAPACK version and are not removed or changed in some incompatible way in some subsequent LAPACK release.
The standard BLAS interface provided by the various existing BLAS libraries is not currently changing, so changes are not generally needed in the wrappers we provide.
Changes to the Cython wrappers for the functions in \texttt{scipy.{\allowbreak}special} follow corresponding changes to the interface of that submodule.

@@

\subsection*{Statistical distributions}
The \texttt{scipy.stats} module provides provides more than 100 probability
distributions and a framework to implement additional distributions. Given
an expression for a probability distribution function (pdf),
the framework will calculate moments, cumulative distribution function (cdf),
differential entropy, and more. Those properties are calculated based on
generic code, e.g., numerical integration of the pdf to obtain the cdf.
Where possible, the generic code is replaced by explicit formulas or more
accurate or efficient implementations. For example, most distributions have
an explicit formula for the cdf, so it is used instead of integrating the
pdf. The distributions framework also allows the user to draw random
variates from the distribution. Drawing random variates now accepts a \texttt{random\_state}
parameter, which is either a \texttt{numpy.random.RandomState} object or an
integer used to generate such an object, to provide fine-grained user-level
control over the random state of a workflow.
Currently about 80 continuous univariate distributions, 12 discrete univariate
distributions and 10 multivariate distributions are implemented.
Key recent distributions added to \texttt{scipy.stats} include the histogram-based
distribution in \texttt{scipy.stats.rv\_histogram} and the multinomial
distribution in \texttt{scipy.stats.multinomial} (used, for example, in natural
language processing\cite{Griffiths5228}). \texttt{multinomial} is the first and so 
far the only multivariate discrete distribution available in SciPy.

@@

While \texttt{scipy.sparse} is mature, there are many data
formats and the community continues to move forward here so we will aim
to take advantage of their improvements\cite{abbasi2018sparse} 
while preserving backwards compatibility to the extent that it is possible.
\texttt{sparse.linalg} may benefit from wrappers for PROPACK 
for faster sparse SVD computation.

@@

\textbf{Maintainers and contributors}
The SciPy project has approximately 15 active maintainers - people who review
others' contributions, and in general do everything needed to ensure that
the software and the project move forward. Maintainers are critical to the
health of the project\cite{eghbal2016}; their skills and efforts largely
determine how fast the project moves forward, and they enable contributions
from a much larger group. Accordingly, the project also has an additional
~100 unique contributors for every 6-monthly release cycle. Anyone with the
interest and skills can become a contributor or maintainer; the SciPy
Developer Guide provides guidance on how to do that.

@@

From the user side, there were 13,096,468 PyPI downloads
of SciPy during the year 2017\cite{pypinfo}, which is likely a vast
underestimate of total downloads given that we don't have access to the
popular Anaconda package manager download statistics. The recommended 
source for citing SciPy is simply a reference to the website / library
itself\cite{SciPylib}, and has been cited $>3000$ times. Some of the most
prominent usages of or demonstrations of credibility for SciPy include the LIGO / Virgo
scientific collaboration that lead to the observation of gravitational
waves \cite{PhysRevLett.116.061102}, and the fact that SciPy is shipped directly
with MacOS and in the Intel Distribution for Python\cite{intel-python}.

\textbf{Future development}.
Future development plans include a major rewrite of the \texttt{scipy.sparse} module, which is already underway\cite{abbasi2018sparse}. This is due to the current interface only supporting two-dimensional objects. Even sub-arrays are two-dimensional, and dimensions greater than two aren't supported. The new package supports dimensions greater than two, and it bases itself around arrays rather than matrices (which is strictly more powerful). It also features improved inter-operability with NumPy and other community packages.

\textit{This section should include key issues: sparse arrays, ndimage pixel vs point, splines, fftpack vs. np.fft and linalg vs. np.linalg, under-maintained submodules.}
